---
title: "Lipidomics data analysis with R"
subtitle: "Aanlyze lipids data generated from Lipidizer"
author: "zhijun cao"
date: "`r Sys.Date()`"
output: 
  html_document:
            number_sections: yes
            theme: journal
            toc: yes
            toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

All files resulted from this analysis were located:
C:/zhijuncao/R/demo/
```{r}
#rm(list=ls())
filepath <- "C:/zhijuncao/R/demo/"
```


```{r loading library}
library(ggplot2)
library(ggpubr)
library(tibble)
library(openxlsx)
library(purrr)
library(cowplot)
library(magrittr)
library(dplyr)
library(tidyr)
library(DT)
```

# Data reshape

## Read lipids MarkerLynx XS Marker Report 

```{r , echo=TRUE}
intest <- read.xlsx("C:/zhijuncao/R/demo/lipids.xlsx", sheet = "Sheet1")

datatable(intest, options = list(scrollX = TRUE))

```

## Reshape function: tranform the lipid report to a tyde table (samples in rows, lipids in columns) 
```{r}
lipid_report_clean <-
  function(tem = intest,
           sample_info = c(1:3),
           lipid_info = c(1:6)) {
    #samples meta data
    sample_meta <- t(tem[sample_info, -lipid_info]) %>% data.frame()
    names(sample_meta) <- c("LCMS", "Animalid", "Samples")
    
    #lipids meta data
    lipids_meta <- tem[-sample_info, lipid_info]
    is_class <- is.na(lipids_meta) %>% rowSums(.)
    
    #create lipid class column
    Class <- c()
    for (i in 1:length(is_class)) {
      if (is_class[i])
      {
        tem_class <- lipids_meta[[1]][i]
      }
      Class[i] <- tem_class
    }
    lipid_meta1 <-
      lipids_meta %>% add_column(uniID = paste0("X", 1:nrow(.)),
                                 Class = Class,
                                 .before = 1)
    names(lipid_meta1)[-c(1:2)] <-
      paste(tem[max(sample_info), lipid_info])
    
    #transpose lipids from row to column
    data_df <-
      t(tem[-sample_info, -lipid_info]) %>% data.frame() %>% mutate_all(as.numeric)
    names(data_df) <- lipid_meta1$uniID
    
    #combined sample meta data with data
    data <- cbind(sample_meta, data_df)
    
    return (list(data = data, ids = lipid_meta1))
    
}
```

## Reshape the report, save the data as excel file
```{r}
intest_clean <- lipid_report_clean(tem=intest, sample_info=c(1:3), lipid_info=c(1:6))
write.xlsx(intest_clean, paste(filepath,"intest_clean.xlsx", sep=""))
```
### Show data, and lipid meta data
```{r}
datatable(intest_clean$data, options = list(scrollX = TRUE))

datatable(intest_clean$ids, options = list(scrollX = TRUE))
```

## Add samples meta data: Type and Day
```{r}
intest_data <- intest_clean$data %>% add_column(Type=substr(intest_clean$data$Samples,1,2),Day=substr(intest_clean$data$Samples, 3,4), .after = 3)

```
```{r}
datatable(intest_data, options = list(scrollX = TRUE))
```


## Functions to deal with 0 or miss values
```{r}
### replace 0 or missing value with 0.5 minimum
is.zero <- Vectorize(
  function(x){
     if (!is.na(x)) {if (x==0) return (TRUE)}
     return (FALSE)
  }
)

###count NA and 0 
countNA0 <- function(x) {
  a <- sum(is.na(x))
  b <- sum(is.zero(x))
  return (a + b)
}
### remove lipid with # of missing
rmNA <- function(data = intest_data, nonNA = 0.5) {
  sumNA <- data %>% map_df( ~ countNA0(.x))
  return (data[, (sumNA / nrow(data)) <= nonNA])
}

### replace missing value with n minimum
repNA_nmin <- function(x, n) {
  #x is vector or list
  if (countNA0(x) != length(x))
    x[is.na(x)] <- min(x, na.rm = TRUE) * n
  return (x)
}

rep0_nmin <- function(x, n) {
  #x is vector or list
  if (countNA0(x) != length(x))
    x[is.zero(x)] <- min(x[x > 0], na.rm = TRUE) * n
  return (x)
}

```
## Replace missing value, 0 values and log transform
```{r}
intest05min <- intest_data %>% rmNA %>% mutate_at(-c(1:5), ~rep0_nmin(.,0.5))
intest05minlog <- intest05min %>% mutate_if(is.numeric, log)

```

## Data summary: mean±sd
### Function of summary
```{r}
data_summary <- function(data = intest05min, ids = intest_clean$ids) {
  summary <- data %>% group_by(Type, Day) %>%
    summarise_if(is.numeric, list(mean = mean, sd = sd)) %>%
    pivot_longer(-c(1:2), names_to = "summary", values_to = "value") %>%
    separate(summary, into = c("uniID", "meansd"), sep = "_") %>%
    pivot_wider(
      id_cols = c(1:3),
      names_from = meansd,
      values_from = value
    ) %>%
    mutate(meansd = paste(round(mean, 2), round(sd, 2), sep = " ± ")) %>% 
    unite("Type_Day", Type, Day) %>%
    pivot_wider(id_cols = uniID,
                names_from = Type_Day,
                values_from = meansd) %>% inner_join(ids, .)
  
  return (summary)
}
```
### Summary data and save as excel file
```{r}
intest05min_summary <- data_summary(data=intest05min, ids=intest_clean$ids)
write.xlsx(intest05min_summary, paste0(filepath,"intest05min_summary.xlsx"))
```
```{r}
datatable(intest05min_summary, options = list(scrollX = TRUE))
```

# Correlation analysis

## Functions of correlation
```{r}
### return coefficient and p value for correlation test
cor_p <- function(...) {
  re <- cor.test(...)
  cor_p <- c(re$estimate, p = re$p.value)
  return (cor_p)
}

### return coefficients and p values as dataframe
cor_p_df <-
  function(data = intest05min,
           x = "X2",
           Y = c("X3", "X4"),
           ids = intest_clean$ids[, c(1:3)]) {
    df <-
      map2_df(x, Y, ~ cor_p(data[[.x]], data[[.y]], method = "pearson")) %>%
      add_column(uniID = Y, .before = 1) %>% inner_join(ids, .)
    return (df)
  }

###all, WT, KO correlation
wt_ko_cor_p <-
  function (data = intest05min,
            x = "X2",
            Y = c("X3", "4") ,
            ids = intest_clean$ids) {
    data1 <- data %>% filter(Type == "WT")
    data2 <- data %>% filter(Type == "KO")
    all <- cor_p_df(data = data,
                    x = x,
                    Y = Y,
                    ids = ids)
    WT <- cor_p_df(data = data1,
                   x = x,
                   Y = Y,
                   ids = ids)
    KO <- cor_p_df(data = data2,
                   x = x,
                   Y = Y,
                   ids = ids)
    df <-
      full_join(WT,
                KO,
                by = c("uniID", "Class", "ID"),
                suffix = c("_WT", "_KO")) %>%
      full_join(all, by = c("uniID", "Class", "ID"))
    return (df)
  }
```
## Correlation: lipids vs cefoperazon
```{r}
intest_cefoperazone <- wt_ko_cor_p(
  data = intest05min,
  x = intest_clean$ids$uniID[intest_clean$ids$ID == "Cefoperazone"],
  Y = names(intest05min)[-c(1:5)],
  ids = intest_clean$ids[, c(1:3)]
)
intest_cefoperazone_sig <-
  intest_cefoperazone %>% filter(p_WT < 0.05 | p_KO < 0.05 | p < 0.05)
write.xlsx(intest_cefoperazone_sig,
           paste0(filepath, "intest_cefoperazone_sig.xlsx"))
```
```{r}
datatable(intest_cefoperazone, options = list(scrollX = TRUE))
```

# Correlation plot
## Function of correlation plot
```{r}
cor_plot <-
  function(data = intest05min,
           xvariable = "Cefoperazone",
           Y = intest_cefoperazone_sig$uniID,
           ids = intest_clean$ids) {
    x <- ids$uniID[ids$ID == xvariable]
    for (y in Y) {
      ylab <- ids$ID[ids$uniID == y]
      #cat(y, " ")
      topy <- max(data[[y]], na.rm = T)
      p1 <- ggscatter(data,
                      x = x,
                      y = y,
                      add = "reg.line") +
        facet_wrap( ~ Type, ncol = 2, scales = "fixed") +
        stat_cor(label.y = topy * 0.82) +
        stat_regline_equation(label.y = topy * 0.9) +
        labs(x = xvariable, y = ylab)
      
      p2 <- ggscatter(data,
                      x = x,
                      y = y,
                      add = "reg.line") +
        facet_wrap( ~ "KO+WT") +
        stat_cor(label.y = topy * 0.82) +
        stat_regline_equation(label.y = topy * 0.9) +
        labs(x = xvariable, y = ylab)
      
      #grid.arrange(p2, p1,ncol=2)
      print(plot_grid(
        p2,
        p1,
        align = "h",
        nrow = 1,
        rel_widths = c(1.1 / 3,  1.9 / 3)
      ))
    }
  }
```
## Plot the correction and save plot as pdf file

### Show examples
```{r}
cor_plot(data=intest05min, xvariable="Cefoperazone",Y=intest_cefoperazone_sig$uniID[1:2], ids=intest_clean$ids)
```

### save all plots to a pdf file
```{r}
#intest Cefoperazone
pdf(paste0(filepath, "intestine_Cefoperazone_corplot_sig.pdf"), width = 8, height = 4)

cor_plot(data=intest05min, xvariable="Cefoperazone",Y=intest_cefoperazone_sig$uniID, ids=intest_clean$ids)

dev.off()
```

# Statistic analysis: linear regression
```{r}
#########################
library(lme4)
library(modelbased)
```


## Group comparison

Compare KO to WT
```{r}
intest05minlog$Type <- factor(intest05minlog$Type, levels = c("WT", "KO"))
```

```{r}

contrasts1 <- c(
  "KO D0 - KO D1",
  "KO D0 - KO D3",
  "KO D0 - KO D5",
  "WT D0 - WT D1",
  "WT D0 - WT D3",
  "WT D0 - WT D5",
  "WT D0 - KO D0",
  "WT D1 - KO D1",
  "WT D3 - KO D3",
  "WT D5 - KO D5"
)

```

## Functions of linear regression analysis

### Function to calculate FDR
```{r}
fdr <- function(x)p.adjust(x, method="BH") # calculate FDR
```

### Function of fixed effect
```{r}

lmfixed <-
  function(data = intest05minlog,
           ids = intest_clean$ids,
           contrasts1 = contrasts1) {
    re <- list()
    ratio <- list()
    for (x in names(data)[-c(1:5)]) {
      #cat(x, "  ")
      fml <- paste(x, "~ Type*Day")
      fml <-  as.formula(fml)
      fit <- lm(fml, data = data)
      
      results <-
        estimate_contrasts(fit, adjust = "none") %>% data.frame()
      
      row.names(results) <-
        paste(results$Level1, results$Level2, sep = " - ")
      
      results1 <- results[contrasts1,]
      re[[x]]  <- p.adjust(results1$p, method = "BH")
      ratio[[x]] <- 1 / exp(results1$Difference)
    }
    
    re1 <- do.call(rbind, re) %>% data.frame()
    ratio1 <- do.call(rbind, ratio) %>% data.frame()
    names(re1) <- contrasts1
    names(ratio1) <- paste(contrasts1, "Ratio", sep = ": ")
    re_fdr <- re1 %>% mutate_all(fdr)
    
    names(re_fdr) <- paste(names(re_fdr), "FDR", sep = ": ")
    names(re1) <- paste(names(re1), "p", sep = ": ")
    pfdr <- cbind(re1, re_fdr, ratio1) %>% 
      tibble::rownames_to_column("uniID")
    
    pfdr1 <- inner_join(ids, pfdr)
    
    return(pfdr1)
  }
```

### Function of mixed effect
```{r}
lmmixed <-
  function(data = stool05minlog,
           ids = stool_clean$ids,
           contrasts1 = contrasts1) {
    re <- list()
    ratio <- list()
    for (x in names(data)[-c(1:6)]) {
      #cat(x, "  ")
      fml <- paste(x, "~ Type*Day + (1|AnimalID)")
      fml <-  as.formula(fml)
      fit <- glmer(fml, data = data)
      
      results <-
        estimate_contrasts(fit, adjust = "none") %>% data.frame()
      row.names(results) <-
        paste(results$Level1, results$Level2, sep = " - ")
      results1 <- results[contrasts1,]
      re[[x]]  <- p.adjust(results1$p, method = "BH")
      ratio[[x]] <- 1 / exp(results1$Difference)
    }
    
    re1 <- do.call(rbind, re) %>% data.frame()
    ratio1 <- do.call(rbind, ratio) %>% data.frame()
    names(re1) <- contrasts1
    names(ratio1) <- paste(contrasts1, "Ratio", sep = ": ")
    re_fdr <- re1 %>% mutate_all(fdr)
    
    names(re_fdr) <- paste(names(re_fdr), "FDR", sep = ": ")
    names(re1) <- paste(names(re1), "p", sep = ": ")
    pfdr <-
      cbind(re1, re_fdr, ratio1) %>% tibble::rownames_to_column("uniID")
    
    pfdr1 <- inner_join(ids, pfdr)
    
    return(pfdr1)
  }
```

## Fixed effect analysis and save result as excel file
```{r}
intest_re <- lmfixed(data = intest05minlog,
                     ids = intest_clean$ids,
                     contrasts1 = contrasts1)
```
Save result to excel file
```{r}
write.xlsx(intest_re, paste0(filepath,"intest_lm_stat.xlsx"))
```
Show first 5 rows
```{r}
datatable(intest_re, options = list(scrollX = TRUE))
```

# Volcanoplot
## Function of volcanoplot
```{r}
library(ggrepel)
volcanoplotfdr <-
  function(pfc,
           title = "volcanoplot",
           fccutoff = 2,
           label = "labelsig",
           labelsize = 3,
           pointsize = 1,
           pcutoff = 0.05,
           fdrcutoff = 2) {
    names(pfc)[1:4] <- c("analytes", "pvalue", "ratio", "fdr")
    
    lowestp <- min(pfc$pvalue[pfc$pvalue > 0])
    pfc$pvalue[pfc$pvalue == 0] <- min(lowestp, 1e-30)
    
    pfc[, "colgroup"] <- "black"
    pfdrcut <- pfc$pvalue  < pcutoff & pfc$fdr < fdrcutoff
    pfdrratio_p <-  pfdrcut & pfc$ratio >= fccutoff
    pfdrratio_n <-  pfdrcut & pfc$ratio <= 1 / fccutoff
    
    pfc[pfdrcut, "colgroup"] <- "orange"
    pfc[pfdrratio_p, "colgroup"] <- "red"
    pfc[pfdrratio_n, "colgroup"] <- "green"
    pfc$colgroup <- as.factor(pfc$colgroup)
    
    changed <-
      c(
        total = nrow(pfc),
        increase = nrow(pfc[pfc$colgroup == "red", ]),
        decrease = nrow(pfc[pfc$colgroup == "green", ])
      )
    
    changed <- paste(changed, collapse = ":")
    
    
    p <- ggplot(data = pfc, aes(log(ratio, 2),-log(pvalue, 10))) +
      theme_bw() +
      geom_point(
        size = pointsize,
        alpha = ifelse(pfc$colgroup == "black", 1, 1),
        color = pfc$colgroup
      ) +
      geom_hline(yintercept = -log10(pcutoff),
                 lty = 3,
                 lwd = 0.5) +
      geom_vline(xintercept = log2(fccutoff) ,
                 lty = 3,
                 lwd = 0.5) +
      geom_vline(xintercept = log2(1 / fccutoff),
                 lty = 3,
                 lwd = 0.5) +
      labs(x = "log2(Ratio)", y = "-log10(p value)") +
      ggtitle(paste(title, changed)) +
      theme(plot.title = element_text(
        color = "black",
        face = "bold",
        size = 12,
        hjust = 0
      )) +
      theme(axis.title = element_text(
        color = "black",
        face = "bold",
        size = 12
      )) +
      theme(plot.margin = unit(c(2, 2, 2, 2), "line"))
    if (label == "labelsig") {
      p + geom_text_repel(
        aes(label = substr(analytes, 1, 25)),
        segment.size = 0.06,
        min.segment.length = 0,
        max.overlaps = 100,
        data = subset(pfc, colgroup == "red" |
                        colgroup == "green"),
        size = labelsize
      )
      
    } else if (label == "nolabel")
    {
      p
    } else if (label == "labelall") {
      p + geom_text_repel(aes(label = analytes), size = labelsize)
      
    }
  }
```

## Read the statistical results
```{r}
intestpfdr <- read.xlsx(paste0(filepath,"intest_lm_stat.xlsx"))
```

## Save  plot to a pdf file
```{r}
tem <- intestpfdr
pdf(paste0(filepath,"intestine_volcanoplot.pdf"), width = 7, height=6)
### KO
pfc <- tem[, c("ID", "KO.D0.-.KO.D1:.p", "KO.D0.-.KO.D1:.Ratio", "KO.D0.-.KO.D1:.FDR")]
p1 <- volcanoplotfdr(pfc,title="KO.D0.-.KO.D1",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p1

pfc <- tem[, c("ID", "KO.D0.-.KO.D3:.p", "KO.D0.-.KO.D3:.Ratio", "KO.D0.-.KO.D3:.FDR")]
p2 <- volcanoplotfdr(pfc,title="KO.D0.-.KO.D3",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p2

pfc <- tem[, c("ID", "KO.D0.-.KO.D5:.p", "KO.D0.-.KO.D5:.Ratio", "KO.D0.-.KO.D5:.FDR")]
p3 <- volcanoplotfdr(pfc,title="KO.D0.-.KO.D5",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p3

#### WT
pfc <- tem[, c("ID", "WT.D0.-.WT.D1:.p", "WT.D0.-.WT.D1:.Ratio", "WT.D0.-.WT.D1:.FDR")]
p4<- volcanoplotfdr(pfc,title="WT.D0.-.WT.D1",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p4

pfc <- tem[, c("ID", "WT.D0.-.WT.D3:.p", "WT.D0.-.WT.D3:.Ratio", "WT.D0.-.WT.D3:.FDR")]
p5 <- volcanoplotfdr(pfc,title="WT.D0.-.WT.D3",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p5
pfc <- tem[, c("ID", "WT.D0.-.WT.D5:.p", "WT.D0.-.WT.D5:.Ratio", "WT.D0.-.WT.D5:.FDR")]
p6 <- volcanoplotfdr(pfc,title="WT.D0.-.WT.D5",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p6

#### WT vs KO
pfc <- tem[, c("ID", "WT.D0.-.KO.D0:.p", "WT.D0.-.KO.D0:.Ratio", "WT.D0.-.KO.D0:.FDR")]
p7 <- volcanoplotfdr(pfc,title="WT.D0.-.KO.D0",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p7

pfc <- tem[, c("ID", "WT.D1.-.KO.D1:.p", "WT.D1.-.KO.D1:.Ratio", "WT.D1.-.KO.D1:.FDR")]
p8 <- volcanoplotfdr(pfc,title="WT.D1.-.KO.D1",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p8

pfc <- tem[, c("ID", "WT.D3.-.KO.D3:.p", "WT.D3.-.KO.D3:.Ratio", "WT.D3.-.KO.D3:.FDR")]
p9 <- volcanoplotfdr(pfc,title="WT.D3.-.KO.D3",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p9

pfc <- tem[, c("ID", "WT.D5.-.KO.D5:.p", "WT.D5.-.KO.D5:.Ratio", "WT.D5.-.KO.D5:.FDR")]
p10 <- volcanoplotfdr(pfc,title="WT.D5.-.KO.D5",fccutoff=1.2, label="labelsig",labelsize=3,pointsize=1, pcutoff=0.05, fdrcutoff=2)
p10
dev.off()
```
## Show volcanoplots
```{r}
p1
p2
p3
p4
p5
p6
p7
p8
p9
p10
```


# Point errobar plots
## Function of point errorbar plots
```{r}
point_errorbar_plot <-
  function(data = intest05min,
           analytes = names(intest05min)[c(6:7)],
           ids = intest_clean$ids) {
    for (analyte in analytes) {
      title <- ids$ID[ids$uniID == analyte]
      p <- ggplot(data, aes(Day, get(analyte))) +
        facet_wrap( ~ Type) +
        theme_classic() +
        geom_point(alpha = 0.5) +
        stat_summary(
          fun.data = "mean_sdl",
          fun.args = list(mult = 1),
          geom = "errorbar",
          width = 0.4,
          position = position_dodge(0.6)
        ) +
        stat_summary(
          aes(y = get(analyte)),
          fun = "mean",
          geom = "point",
          shape = 4,
          position = position_dodge(0.6)
        ) +
        labs(y = "Intensity", x = "Day", title = title) +
        theme(
          axis.title = element_text(size = 10),
          axis.text = element_text(size = 10),
          axis.text.x = element_text(angle = 0)
        ) +
        theme(plot.title = element_text(size = 7))
      
      print(p)
    }
    
  }
```

## Plot point errorbar plots

Show examples
```{r}
point_errorbar_plot(data = intest05min,
                    analytes = names(intest05min)[-c(1:5)][1:2],
                    ids = intest_clean$ids)
```

Save plots to pdf file
```{r}
pdf(paste0(filepath,"intest05min_errorbar.pdf"),width=3.5,height=2.6)
point_errorbar_plot(data = intest05min,
                    analytes = names(intest05min)[-c(1:5)],
                    ids = intest_clean$ids)
dev.off()
```

# Heat map
```{r}
library(gplots)
library(RColorBrewer)
```

## Prepare data for heatmap
```{r}
colors <- brewer.pal(9, "Set1")
names(colors) <- unique(intest_clean$ids$Class)
toMean <- function(x) x/mean(x, na.rm=TRUE)

prepheatmap <- function(data=intest05min, ids=intest_clean$ids){
  
  labrow <- left_join(data.frame(uniID=names(data)[-c(1:5)]),ids)
  toMean1 <- data[data$Type=="KO" ,] %>% mutate_at(-c(1:5),list(toMean))
  toMean2 <- data[data$Type=="WT" ,] %>% mutate_at(-c(1:5),list(toMean))
  toMean12 <- rbind(toMean1,toMean2)
  
  return (list(data=toMean12, labrow=labrow))
}
```

## Save heatmap to pdf file
```{r, fig.width=11, fig.height=8}
intest_hp <- prepheatmap(data=intest05min, ids=intest_clean$ids)
toMean12 <- intest_hp$data
labrow <- intest_hp$labrow
names(colors) <- unique(intest_clean$ids$Class)
#pdf(paste0(filepath, "intestine_KO(tomean)_WT(tomean)_heatmap_color.pdf"), width = 10,height = 10)
heatmap.2(as.matrix(t(log(toMean12[,-c(1:5)]))),col=bluered, Rowv = F, Colv = FALSE, margins = c(4, 15),
          labCol=toMean12$Sample,labRow = labrow$ID ,trace="none",
          dendrogram="none",
          lhei = c(1,6),
          main = "Intestine",
          RowSideColors=colors[labrow$Class], 
          na.color = "grey",srtCol=90, cexCol = 0.8, cexRow = 0.5, key.xlab = "log(toMean)",
          density.info="none",breaks = seq(-1, 1, 0.1))
legend("left", legend = names(colors)[-9],fill = colors,title = "Class", cex = 0.7)
#dev.off()

```






