---
title: "Data analysis"
author: "Zhijun Cao"
date: "7/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message =  TRUE)
```

## Descfription

This document provides R source code for data analyis of the submitted manuscript (prostatic metabolomics)


```{r}
library(openxlsx)
library(mixOmics)
library(tidyr)
library(dplyr)
library(magrittr)
library(tibble)
library(sva)
library(gridExtra)
library(ggplot2)
library(plotly)
library(purrr)
library(multcomp)
```

## Import data
```{r}
lipid <- read.xlsx("lipid_sample-02132020-2.xlsx")
aa <- read.xlsx("AA_sample-02132020.xlsx")
```
## PCA analysis to identify batch effects
```{r}
lipid_pca <- pca(log(lipid[,-(1:6)]), scale=T, ncomp=2)
aa_pca <- pca(log(aa[,-(1:6)]), scale=T, ncomp=2)
```

```{r}
p1 <- plotIndiv(lipid_pca, group=lipid$Batch, pch=as.factor(lipid$Group), legend = TRUE, ind.names=F, title = "lipids: before correction")
p2 <- plotIndiv(aa_pca,  group=aa$Batch, pch=as.factor(aa$Group), legend = TRUE, ind.names=F, title = "AAs: before correction")
```

## Empirical Bayes method for batch correction
### lipids
```{r}
lipid_pheno <- lipid[,c(1:6)]
lipid_data  <-  log(t(as.matrix(lipid[,-c(1:6)])))
colnames(lipid_data) <- lipid_pheno$SampleID
lipid_mod = model.matrix(~as.factor(Group), data=lipid_pheno)

lipid_combat = ComBat(dat=lipid_data, batch=lipid_pheno$Batch, mod=lipid_mod, par.prior=TRUE, ref.batch="b2", prior.plots=FALSE)
lipid_combat_df <- cbind(lipid_pheno, data.frame(t(lipid_combat)))

write.xlsx(lipid_combat_df, "loglipid_combat.xlsx")
```

### AAs
```{r}
aa_pheno <- aa[,c(1:6)]
aa_data  <-  log(t(as.matrix(aa[,-c(1:6)])))
colnames(aa_data) <- aa_pheno$SampleID
aa_mod = model.matrix(~as.factor(Group), data=aa_pheno)

aa_combat = ComBat(dat=aa_data, batch=aa_pheno$Batch, mod=aa_mod, par.prior=TRUE, ref.batch="b2", prior.plots=FALSE)
aa_combat_df <- cbind(aa_pheno, data.frame(t(aa_combat)))

write.xlsx(aa_combat_df, "logaa_combat.xlsx")
```

### PCA analysis after batch correction
```{r}
lipid_combat_df_pca <- pca(lipid_combat_df[,-(1:6)], scale=T, ncomp=2)
aa_combat_df_pca <- pca(aa_combat_df[,-(1:6)], scale=T, ncomp=2)
```

```{r}
p3 <- plotIndiv(lipid_combat_df_pca, group=lipid$Batch, pch=as.factor(lipid$Group), legend = TRUE, ind.names=F, title = "lipids: after correction")
p4 <- plotIndiv(aa_combat_df_pca,  group=aa$Batch, pch=as.factor(aa$Group), legend = TRUE, ind.names=F, title = "AAs: after correction")
```

### PCA plots before and after batch correction
```{r}
pdf("Before and after batch correction.pdf", width = 16, height = 11)
grid.arrange(p1$graph, p3$graph, p2$graph, p4$graph, nrow=2)
dev.off()
```

## PCA and PLSDA analysis for each time point

###function for pca analysis
```{r}
pca_analysis <- function(title="PND30", tem=tem, colors=colors){
  
tem_pca <- pca(tem[,-(1:6)], scale = TRUE, ncomp = 3)

write.xlsx(data.frame(tem_pca$explained_variance), paste(title," pca results.xlsx"), rowNames=T)  
  
  
#3d pca

p <- plot_ly(data.frame(tem_pca$variates$X), x = ~PC1, y = ~PC2, z = ~PC3, color = ~tem$Treatment, colors = colors) %>%
 add_markers() %>%
  layout(scene = list(xaxis = list(title = 'PC1'),
                      yaxis = list(title = 'PC2'),
                      zaxis = list(title = 'PC3')))
p
htmlwidgets::saveWidget(as_widget(p), paste(title," pca_3d.html", sep=""))

#variable correlation and loading
pdf(paste(title, "pca_correlation and loading.pdf"),width=14, height=14)
plotVar(tem_pca, comp=c(1,2), cex=4)
plotVar(tem_pca, comp=c(1,3), cex=4)
plotVar(tem_pca, comp=c(2,3), cex=4)

plotLoadings(tem_pca,comp=1, size.name = 0.3, ndisplay = 100)
plotLoadings(tem_pca,comp=2, size.name = 0.3, ndisplay = 100)
plotLoadings(tem_pca,comp=3, size.name = 0.3, ndisplay = 100)

dev.off()

}
```

###function for plsda analysis
```{r}
plsda_analysis <- function(title="PND30", tem=tem, colors=colors){
  
tem_plsda <- plsda(tem[,-(1:6)],tem$Group, scale = TRUE, ncomp = 3)
perf_plsda <- perf(tem_plsda, validation = "loo", progressBar = FALSE, auc = TRUE)
write.xlsx(list(explained=data.frame(tem.plsda$explained_variance),
           validatonerror=perf_plsda$error.rate$BER), paste(title," plsda results.xlsx"), rowNames=T)  
  
  
#3d plsda

p <- plot_ly(data.frame(tem_plsda$variates$X), x = ~comp1, y = ~comp2, z = ~comp3, color = ~tem$Treatment, colors = colors) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'X−variate 1'),
                      yaxis = list(title = 'X−variate 2'),
                      zaxis = list(title = 'X−variate 3')))
p
htmlwidgets::saveWidget(as_widget(p), paste(title," plsda_3d.html", sep=""))

#variable correlation and loading
pdf(paste(title, "plsda_correlation and loading.pdf"), width=14, height=14)
plotVar(tem_plsda, comp=c(1,2), cex=4)
plotVar(tem_plsda, comp=c(1,3), cex=4)
plotVar(tem_plsda, comp=c(2,3), cex=4)

plotLoadings(tem_plsda,comp=1,method='median', contrib="max", size.legend = 0.5, size.name = 0.3, ndisplay = 100 )
plotLoadings(tem_plsda,comp=2,method='median', contrib="max", size.legend = 0.5, size.name = 0.3, ndisplay = 100 )
plotLoadings(tem_plsda,comp=3,method='median', contrib="max", size.legend = 0.5, size.name = 0.3, ndisplay = 100 )
dev.off()
}
```

####PCA and PLSDA analysis for lipids

```{r}
lipid_combat_df <- lipid_combat_df %>% filter(Animal.ID!=583)
```

```{r}
tem <- lipid_combat_df %>% filter(PND=="PND 30")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB-treated'))
colors <- c('black', 'red')
plsda_analysis(title="Lipid PND30", tem=tem, colors=colors)
pca_analysis(title="Lipid PND30", tem=tem, colors=colors)
```

```{r}
tem <- lipid_combat_df %>%  filter(PND=="PND 90")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB-treated'))
colors <- c('black', 'red')
plsda_analysis(title="Lipid PND90", tem=tem, colors=colors)
pca_analysis(title="Lipid PND90", tem=tem, colors=colors)
```

```{r}
tem <- lipid_combat_df %>% filter(PND=="PND 100")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB only', 'T+E only', 'EB+T+E'))
colors <- c('black', 'red', 'green', 'purple')
plsda_analysis(title="Lipid PND100", tem=tem, colors=colors)
pca_analysis(title="Lipid PND100", tem=tem, colors=colors)
```

```{r}
tem <- lipid_combat_df %>% filter(PND=="PND 145")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB only', 'T+E only', 'EB+T+E'))
colors <- c('black', 'red', 'green', 'purple')
plsda_analysis(title="Lipid PND145", tem=tem, colors=colors)
pca_analysis(title="Lipid PND145", tem=tem, colors=colors)
```

```{r}
tem <- lipid_combat_df %>% filter(PND=="PND 200")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB only', 'T+E only', 'EB+T+E'))
colors <- c('black', 'red', 'green', 'purple')
plsda_analysis(title="Lipid PND200", tem=tem, colors=colors)
pca_analysis(title="Lipid PND200", tem=tem, colors=colors)
```

####PCA and PLSDA analysis for amino acids
```{r}
tem <- aa_combat_df %>% filter(PND=="PND 30")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB-treated'))
colors <- c('black', 'red')
plsda_analysis(title="AA PND30", tem=tem, colors=colors)
pca_analysis(title="AA PND30", tem=tem, colors=colors)
```

```{r}
tem <- aa_combat_df %>% filter(PND=="PND 90")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB-treated'))
colors <- c('black', 'red')
plsda_analysis(title="AA PND90", tem=tem, colors=colors)
pca_analysis(title="AA PND90", tem=tem, colors=colors)
```

```{r}
tem <- aa_combat_df %>% filter(PND=="PND 100")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB only', 'T+E only', 'EB+T+E'))
colors <- c('black', 'red', 'green', 'purple')
plsda_analysis(title="AA PND100", tem=tem, colors=colors)
pca_analysis(title="AA PND100", tem=tem, colors=colors)
```

```{r}
tem <- aa_combat_df %>% filter(PND=="PND 145")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB only', 'T+E only', 'EB+T+E'))
colors <- c('black', 'red', 'green', 'purple')
plsda_analysis(title="AA PND145", tem=tem, colors=colors)
pca_analysis(title="AA PND145", tem=tem, colors=colors)
```

```{r}
tem <- aa_combat_df %>% filter(PND=="PND 200")
tem$Treatment <- factor(tem$Treatment, levels = c('CTRL', 'EB only', 'T+E only', 'EB+T+E'))
colors <- c('black', 'red', 'green', 'purple')
plsda_analysis(title="AA PND200", tem=tem, colors=colors)
pca_analysis(title="AA PND200", tem=tem, colors=colors)
```

## Boxplot

```{r}
Boxplots <- function(pdfname="loglipid_combat_boxplot_fill.pdf", data=data){
  level=c( "PND 30 CTRL", "PND 30 EB-treated",
          "PND 90 CTRL", "PND 90 EB-treated",
          "PND 100 CTRL", "PND 100 EB only", "PND 100 T+E only", "PND 100 EB+T+E", 
          "PND 145 CTRL", "PND 145 EB only",  "PND 145 T+E only", "PND 145 EB+T+E",  
          "PND 200 CTRL", "PND 200 EB only", "PND 200 T+E only", "PND 200 EB+T+E" 
          )
data$Group <-factor(data$Group, levels = level)
data$PND <-factor(data$PND, levels = c("PND 30","PND 90", "PND 100", "PND 145", "PND 200"))
analytes <- names(data)[-c(1:6)]
pdf(pdfname, width = 6, height = 5)
for (analyte in analytes){
  p <- ggplot(data, aes(Group, get(analyte), fill=PND))+
  theme_bw()+ 
  geom_boxplot()+
  labs(y=analyte)+
  theme(axis.title=element_text(size=10),axis.text=element_text(size=10),axis.text.x=element_text(angle=90))+
  theme(plot.title=element_text(size=14))+
  theme(plot.margin=unit(c(1,1,1,1),"line"))
  print(p)
}
dev.off()
}
```

### Lipids boxplot
```{r}
Boxplots(pdfname="lipid_combat_boxplot_fill.pdf", data=lipid_combat_df)
```

### Amino acid boxplot
```{r}
Boxplots(pdfname="aa_combat_boxplot_fill.pdf", data=aa_combat_df)
```

## Statistical analysis
```{r}
aov_dunn <- function(excelname="lipid_combat_ratio_pvalue_fdr.xlsx", data=lipid_combat_df){
  
  Mean <- function(x)mean(x, na.rm=TRUE)
  ratio <- function(x)exp(x-x[1])
  
  data$Treatment <- factor(data$Treatment)
  PND<- unique(data$PND)
  analytes <- names(data)[-c(1:6)]
  pnd_re <- list()

  for (pnd in PND) {
  
    tem <- data %>% filter(PND==pnd)
    ratio_re <- tem %>% select(-Animal.ID)%>% group_by(Group) %>% summarise_if(is.numeric,Mean) %>% 
      mutate_if(is.numeric, ratio)
    ratio_re_df <- data.frame(t(as.matrix(ratio_re[,-1]))) %>% select(-1)
  
    pvalues <- list()
    
    for (x in analytes){
      m1 <- aov(tem[,x] ~ Treatment, data=tem)
      re <- glht(m1, linfct=mcp(Treatment="Dunnett"))
      sum_re_pvalue <- summary(re)$test$pvalues
      names(sum_re_pvalue) <- names(coefficients(re))
      pvalues[[x]] <- sum_re_pvalue
    }
    
    df_re <- data.frame(do.call(rbind, pvalues))
    fdr_re <- map_df(df_re, ~p.adjust(.x, method="BH"))
    names(fdr_re) <- paste(names(df_re), " FDR")
    names(ratio_re_df) <- paste(names(df_re), " Ratio")
    ratio_re_df <- ratio_re_df %>% rownames_to_column(var="analytes")
    names(df_re) <- paste(names(df_re), "p value")
    pnd_re[[pnd]] <- cbind(ratio_re_df, df_re, fdr_re)
}

write.xlsx(pnd_re, excelname ,rownames=TRUE)
}

```

### Lipid stat analysis
```{r}
aov_dunn(excelname="ipid_combat_ratio_pvalue_fdr.xlsx", data=lipid_combat_df)
```

### Amino acid stat analyis
```{r}
aov_dunn(excelname="aa_combat_ratio_pvalue_fdr.xlsx", data=aa_combat_df)
```
